The CMS Data Aggregation System

### 2010-05-04 doc/paper.tex
Incorporate Dave's changes
git-svn-id: svn+ssh://svn.cern.ch/reps/CMSDMWM/DAS/trunk@1844 4525493e-7705-40b1-a816-d608a930855b

\subsection{DAS core\label{DAS_core}}
The DAS core library was responsible for data retrieval on demand. 
It had the following components: input query validator and parser,
query analyzer, workflow builder, execution engine and data-service 
plug-ins. The communication between client (web server or CLI) and 
DAS cache server was done via the REST \cite{REST} interface. 
%For each GET request we performed the following action.
Each input query was recorded into Analytics DB for further 
analysis. It was decomposed into the set of conditions and/or selection keys.
First the results were looked up in a cache and delivered to the requestor
if superset of input conditions was found in analytics DB.\footnote{It is important
to note here that we didn't look-up results in a cache and rather
compared input query condition set to the records in analytics DB 
in order to understand if other API calls are covered the input request.}
Otherwise, the appropriate workflow, a set of API calls, was constructed
for the execution engine. It invoked the API calls via data-service
plug-ins and stored results into the cache. At this step we also
performed necessary data aggregation for records with similar key-value
pairs. The data-service plug-in implementation were similar and their 
logic was abstracted, while a few extentions were required, e.g. specific 
security modules for data-services accessible via private network.
Finally, the results were retrieved from the cache and delivered to the user.

\subsection{DAS Mapping DB}
The Mapping DB represents DAS {\it registration} service. It collects
information about data-service APIs, translates API input/output
parameters into DAS keys and contains mapping between DAS records
and their UI representation.

@@

\subsection{DAS caching system}
The DAS caching system was used to dynamically fetch
and aggregate data upon user requests. We used the following 
algorithm (code snippets are written in Python \cite{Python}):
\begin{enumerate}[1.]
\item Decompose query into set of conditions (\verb+conds+) 
and selection keys (\verb+skeys+).
\item Check cache for query hash. If found, look-up results from the cache,
otherwise proceed to the next step.
\item Identify list of APIs which can serve requested data
\begin{verbatim}
apilist = [r for r in mapping_db(skeys, conds)]
\end{verbatim}
Here \verb+r+ represents a triplet \verb+url, api, args+ and 
\verb+mappind_db+ defines translation between input query and data-service
APIs which can delivery the data.
\item Iterate over APIs
\begin{verbatim}
for url, api, args in apilist:
    if  superset_in_analytics(api, args):
        continue
    for row in parser(get_data(url, api, args)):
        yield row
\end{verbatim}
Here \verb+superset_in_analytics+ checks if provided \verb+api, args+
are subset of already registered in Analytics DB api call(s). The 
\verb+get_data+ and \verb+parser+ are define methods to fetch and parse
data, respectively.
\item Perform aggregation by iterating over records from the previous step
\begin{verbatim}
if cache_has(row.primary_key):
   existing_record = get_from_cache(row.primary_key)
   merge(row, existing_record)
   insert_into_cache(row)
   delete_from_cache(existing_record)
else:
   insert_into_cache(row)
\end{verbatim}
The \verb+primary_key+
is a key which identifies the record defined in Mapping DB, e.g. \verb+file.name+
for \verb+file+ record.
We did use a bulk insert operations at this step to increase throughput. 
\item Get results from the cache
\end{enumerate}
We also had an option to run independent set of daemons, DAS robots, who
monitor Analytics DB for most popular requests and pre-fetch the most
common data into DAS cache.

\subsection{DAS queries}
Apart from the previous DBS-QL syntax we decided to drop its syntax and allow users
express their queries in a free text-based form. Although we must clarify its
meaning. We used a pre-defined set of keys registered in DAS to describe common
entities of our data. For example, \verb+dataset+, 
\verb+file+, \verb+block+, \verb+run+, \verb+event+, etc. Their choice
was based on analysis of daily work operations of our users, see \cite{DBS07}.
Each key had a set of attributes with primary one assigned to the key itself.
For instance, the \verb+block+ key had the following attributes
\verb+name+, \verb+size+, \verb+replica+ etc., 
with primary attribute being \verb+name+.
The choice of attributes were based on a set of meta-data information
associated with a given DAS key. Therefore end-users where able to specify
a context in their queries by using appropriate attribute(s). For example,
for query \verb+block+ we didn't perform a string matching to the
``block'' string, but rather look-up a block objects in a DAS, while query
\verb+block.name=abc*+ instructs system to look-up appropriate blocks
with matched names. The numerical values, conditions, groupping were
easy to express too. For example, \verb+run > 10 and run < 100+ query 
was used to get all run records for provided range of run numbers. 
At the same time, to complement the strength of the 
DAS, we do plan to index dynamically retrieved meta-data and add a keyword-based 
search later.

The free text-based queries were transformed into MongoDB query syntax, 
who was used internally in DAS. It represents a set of selection 
keys and conditions as a JSON document. For example, a query
{\it dataset site=T1\_CH\_CERN}
was transformed into 
{\it \{"fields" : ["dataset"], "spec" : \{"site" : "T1\_CH\_CERN"\}\}}.
Such syntax was perfect fit for us, since it allows to store all queries
as JSON documents into Analytics DB for further analysis. It is worthwhile to note that
MongoDB query syntax is very reach. It supports advanced queries with 
boolean expressions, conditional operators, regular expressions, etc. providing
a real power for our end-users.

\subsection{Data-services and aggregation}
As we outlined above the DAS Mapping DB becomes an authorative
source of information about all data-services participating in DAS.
Based on information stored in Mapping DB we were able to 
retrieve appropriate data from different data-services, re-map them into 
DAS notations and aggregate them on demand. For instance, the Data 
Bookkeeping System (DBS) and data location system (PhEDEx)
provided information about CMS block meta-data. In former case, DBS 
stored information about blocks and their relationships to other 
physics objects, such as dataset and files, while PhEDEx kept
information about its location and replicas. Upon a user request to 
find a block meta-data, we were able to identify appropriate set of DBS and PhEDEx APIs, 
retrieve data from them, re-format their output to DAS notations 
and store results into DAS cache performing aggregation of matched 
objects using \verb+block.name+ key. The merged results were stored 
as a set of data records. Here is an example of such aggregated record:
\begin{verbatim}
{"das_id": ["4b195404e2194e21c4000030", "4b1953fee2194e21c4000002"], 
 "block": [{"name": "/a/b/c#123", "replica": [{"complete": "y", 
            "subscribed": "n", "custodial": "n", "site": "T1_US_Buffer", 
            "nfiles": "18", "se": "srm.site.com", 
            "size": "112694171838"}, {....}]},
           {"name": "a/b/c#123", "nevents": "570200", 
            "created_by": "USER_DN", "path": "/a/b/c", ...}],
 "das": [{"expire": 1259952908}, {"expire": 1259954702}]}
\end{verbatim}
Here \verb+das_id+ refers to api call records stored in Analytics DB,
\verb+block+ represents an aggregation block meta-data information from DBS
and PhEDEx systems and \verb+das+ contains their expiration time stamps.
Usage of this information was valuable contribution in debugging process of
data-services themselves, e.g. identification of not-matched records, 
latency studies, etc.
% Here is an example of DAS record representing
%aggregation information between two SiteDB APIs, {\it CMStoSAMName} 
%and {\it CMStoSiteName}, with API results merged under the {\it site} key.
%\begin{verbatim}
%{"site":[{"samname": "CERN-PROD", "name": "T1_CH_CERN"}, 
%         {"sitename": "CERN", "name": "T1_CH_CERN"}], 
% "das": [{"ctime": 0.9, "expire": 1256362175, "api": ["CMStoSAMName"],
%          "url": "https://a.b.com/sitedb/json/index/CMStoSAMName", 
%          "timestamp": 1256318975, "system": ["sitedb"]},
%         {"ctime": 1.2, "api": ["CMStoSiteName"], "expire": 1256362174},
%          "url": "https://a.b.com/sitedb/json/index/CMStoSiteName", 
%          "timestamp": 1256318974, "system": ["sitedb"]]}
%\end{verbatim}

%We used RESTful model \cite{REST} for all DAS cache server APIs.
%It allows us to share core library and other components during development of 
%various DAS clients, e.g. CLI, web interface, robots, etc. and hides
%complexity of access to individual data-services and their security policies
%as well as proper organize caching solution.
%For instance, to access on-line Run Summary information we were forced to use
%Grid certificates. Instead of propagating user certificate to data-service
%the DAS cache server works as a proxy to this server and used host certificate 
%to serve such requests on behalf of the users.

\section{Results\label{Results}}
In this section we discuss preliminary results achieved with DAS system
using the following CMS data-services:
\begin{itemize}
\item Data Bookkeeping System, DBS \cite{DBS}, the CMS meta-data service;
\item Physics Experiment Data Export, PhEDEx \cite {PhEDEx}, the 
CMS file transfer system;
\item SiteDB \cite{SiteDB}, the CMS site resources data-service;
\item RunSummary \cite{RunSummary}, the CMS run and trigger data-service;
%\item LumiDB \cite{LumiDB} collects information about Luminosity conditions;
%\item DQ \cite{DQ}, Data Quality data-service which collects information
%about detector conditions during data taking;
%\item Overview \cite{Overview}, collects information about CMS
%transfer rate, etc.;
\item Dashboard \cite{Dashboard}, the LHC virtual organization
monitoring data-service portal;
\end{itemize}
Each data-service has its own scope, size and peculiar properties. 
For instance, the data in PhEDEx are transient due to constant 
migration of CMS data; the DBS system were divided into dozen 
of individual instances with own size and lifetime, who has been used by different production 
teams for various bookkeeping tasks; the RunSummary data-service
operates on private network accessible at CMS online operation center.
%Table \ref{db_size} shows the current size of meta-data from aforementioned 
%data-services.
%\begin{figure}[htb]
%\centering
%\includegraphics[width=100mm]{db_size.pdf}
%\caption{
%Current size of meta-data for CMS data-services participated in DAS, before
%official data-taking. 
%}
%\label{db_size}
%\end{figure}

>>>>

\subsection{DAS core\label{DAS_core}}
The DAS core library is responsible for data retrieval on demand. 
It has the following components: input query validator and parser,
query analyzer, workflow builder, execution engine and data-service 
plug-ins. The communication between client (web server or CLI) and 
DAS cache server was done via the REST \cite{REST} interface. 
Each input query was recorded into the Analytics DB for further 
analysis, and was decomposed into a set of conditions and/or selection keys.
First the results were looked up in a cache and delivered to the requestor
if superset of input conditions was found in analytics DB.\footnote{It is important
to note here that we didn't look-up results in a cache and rather
compared input query condition set to the records in analytics DB 
in order to understand if other API calls are covered the input request.}
Otherwise, the appropriate workflow, a set of API calls, was constructed
for the execution engine. It invoked the API calls via data-service
plug-ins and stored results into the cache. At this step we also
performed necessary data aggregation for records with similar key-value
pairs. The data-service plug-in implementations are similar and their 
logic is largely abstracted, apart from a few extentions such as specific
security modules for data-services accessible via private network.
Finally, the results were retrieved from the cache and delivered to the user.

\subsection{DAS Mapping DB\label{MappingDB}}
The Mapping DB represents DAS {\it registration} service. It collects
information about data-service APIs, translates API input/output
parameters into DAS keys and contains mapping between DAS records
and their UI representation.

@@

\subsection{DAS caching system}
The DAS caching system was used to dynamically fetch
and aggregate data upon user requests. We used the following 
algorithm (code snippets are written in Python \cite{Python}):
\begin{enumerate}[1.]
\item Decompose query into set of conditions (\verb+conds+) 
and selection keys (\verb+skeys+).
\item Check cache for query hash. If found, look-up results from the cache,
otherwise proceed to the next step.
\item Identify list of APIs which can serve requested data
\begin{verbatim}
apilist = [r for r in mapping_db(skeys, conds)]
\end{verbatim}
Here \verb+r+ represents a triplet \verb+url, api, args+ and 
\verb+mappind_db+ defines translation between input query and data-service
APIs which can delivery the data.
\item Iterate over APIs
\begin{verbatim}
for url, api, args in apilist:
    if  superset_in_analytics(api, args):
        continue
    for row in parser(get_data(url, api, args)):
        yield row
\end{verbatim}
Here \verb+superset_in_analytics+ checks if provided \verb+api, args+
are subset of already registered in Analytics DB api call(s). The 
\verb+get_data+ and \verb+parser+ are define methods to fetch and parse
data, respectively.
\item Perform aggregation by iterating over records from the previous step
\begin{verbatim}
if cache_has(daskey):
   existing_record = get_from_cache(daskey)
   merge(row, existing_record)
   insert_into_cache(row)
   delete_from_cache(existing_record)
else:
   insert_into_cache(row)
\end{verbatim}
The \verb+daskey+
is a key which identifies the record defined in Mapping DB, e.g. \verb+file.name+
for \verb+file+ record.
We did use a bulk insert operations at this step to increase throughput. 
\item Get results from the cache
\end{enumerate}
We also had an option to run independent set of daemons, DAS robots, who
monitor Analytics DB for most popular requests and pre-fetch the most
common data into the cache.

\subsection{DAS queries}
All DAS queries were expressed in a free text-based form, either as a 
set of keywords or key-value pairs, where a pair can represent
condition, e.g. \verb+key > value+. We used a pre-defined set of keys registered 
in DAS to describe common entities of our data. For example, \verb+dataset+, 
\verb+file+, \verb+block+, \verb+run+, \verb+event+, etc. Their choice
was based on the analysis of daily work operations of our users, see \cite{DBS07}.
Each key represents a DAS record. Every record was stored into the cache
in JSON format. Due to the ``schemaless'' nature of underlying back-end (MongoDB)
we were able to store DAS records of arbitrary structure, e.g.
dictionary, lists, key-value pairs, etc. Therefore every DAS key had a 
set of attributes representing JSON structure. For instance a document
\begin{verbatim}
{"block": {"name": "abc", "size": 1, "replica": [...]}}
\end{verbatim}
associated with a \verb+block+ key who had the following attributes
\verb+name+, \verb+size+, \verb+replica+. The set of attributes is based 
on a set of meta-data information stored in a record and therefore
a nested structures were allowed, e.g. \verb+block.replica.size+. Our users 
where able to specify a context of their queries by using appropriate 
attribute(s). For example, query \verb+block+ represents search for all
block records rather then ``block'' string, while query
\verb+block.name=abc*+ instructs the system to look-up appropriate blocks
with matched names. The numerical values, conditions, groupping were
easy to express too, e.g. \verb+run > 10 and run < 100+ query 
was used to get all run records for provided range of run numbers. 
To complement the strength of the queries, we do plan to index 
dynamically DAS records and add a keyword-based 
search within record content later.

All input queries were transformed into MongoDB query syntax who is itself
a JSON record. It contains a set of selection 
keys (\verb+fields+) and conditions (\verb+spec+). For example, a query
\verb+dataset, file=abc, run>10+
was represented as
\begin{verbatim}
{"fields": ["dataset"], "spec": {"file": "abc", "run": {"$gt": 10}}}
\end{verbatim}
Therefore users queries were stored into Analytics DB similar to other DAS 
records and used for further analysis.
%Such syntax was perfect fit for us, since it allows to store all queries
%as JSON documents into Analytics DB for further analysis. It is worthwhile to note that
%MongoDB query syntax is very reach. It supports advanced queries with 
%boolean expressions, conditional operators, regular expressions, etc. providing
%a real power for our end-users.

\subsection{Data-services and aggregation}
As was shown in section \ref{MappingDB} the DAS Mapping DB became an authorative
source of information about all data-services participating in DAS.
Based on the information stored in Mapping DB we were able to 
retrieve appropriate data from different data-services, re-map them into 
DAS notations and aggregate them on demand. For instance, the Data 
Bookkeeping System (DBS) \cite{DBS} and data location system (PhEDEx) \cite{PhEDEx}
were provided information about CMS blocks. In former case, DBS 
stored information about blocks and their relationships to other 
physics objects, such as dataset and files, while PhEDEx kept
information about its location and replicas. Upon a user request to 
find a block meta-data, we were able to identify appropriate set of 
DBS and PhEDEx APIs, retrieve data, re-format their output 
into DAS notations and store results into DAS cache. The records
were merged (aggregated) based on identical key-value pairs, e.g.
based on \verb+block.name+ key. Here is an example of such aggregated record:
\begin{verbatim}
{"das_id": ["4b195404e2194e21c4000030", "4b1953fee2194e21c4000002"], 
 "block": [{"name": "/a/b/c#123", "replica": [{"complete": "y", 
            "subscribed": "n", "custodial": "n", "site": "T1_US_Buffer", 
            "nfiles": "18", "se": "srm.site.com", 
            "size": "112694171838"}, {....}]},
           {"name": "a/b/c#123", "nevents": "570200", 
            "created_by": "USER_DN", "path": "/a/b/c", ...}],
 "das": [{"expire": 1259952908}, {"expire": 1259954702}]}
\end{verbatim}
The \verb+das_id+ part refers to api call records stored in Analytics DB,
the \verb+block+ part represents an aggregation block meta-data information from DBS
and PhEDEx systems and the \verb+das+ part contains their expiration time stamps.
Usage of this information was valuable contribution in debugging process of
data-services themselves, e.g. identification of data-service inconsistencies, 
latency studies, etc.

\section{Results\label{Results}}
In this section we discuss preliminary results achieved with DAS system
using the following CMS data-services:
\begin{itemize}
\item Data Bookkeeping System, DBS \cite{DBS}, the CMS meta-data service;
\item Physics Experiment Data Export, PhEDEx \cite {PhEDEx}, the 
CMS data location and file transfer system;
\item SiteDB \cite{SiteDB}, the CMS site resources;
\item RunSummary \cite{RunSummary}, the CMS run and trigger data-service;
%\item LumiDB \cite{LumiDB} collects information about Luminosity conditions;
%\item DQ \cite{DQ}, Data Quality data-service which collects information
%about detector conditions during data taking;
%\item Overview \cite{Overview}, collects information about CMS
%transfer rate, etc.;
\item Dashboard \cite{Dashboard}, the LHC virtual organization
monitoring data-service portal;
\end{itemize}
Each data-service had its own scope, size and peculiar properties. 
For instance, the data in PhEDEx were transient due to constant 
migration of CMS data; the DBS system was divided into dozen 
of individual services (instances) with own size and lifetime 
used by different production teams for various bookkeeping tasks; 
the RunSummary data-service operated on private network accessible 
at CMS online operation center.
%Table \ref{db_size} shows the current size of meta-data from aforementioned 
%data-services.
%\begin{figure}[htb]
%\centering
%\includegraphics[width=100mm]{db_size.pdf}
%\caption{
%Current size of meta-data for CMS data-services participated in DAS, before
%official data-taking. 
%}
%\label{db_size}
%\end{figure}
